---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---


## Load packages

```{r}
library(readr)
library(dplyr)
library(skimr)
library(ggplot2)
library(lubridate)
```


## Loading and preprocessing the data

```{r}

activity <- read.csv("activity.csv")
head(activity)
tail(activity)
summary(activity)
str(activity)

skim(activity)
glimpse(activity)

# create a dataset without NAs in case
activity_clean <- na.omit(activity)
summary(activity_clean)

# Convert data column to dates
summary(activity)
activity$date <- as.Date(activity$date)
summary(activity)

year(as.Date(activity$date[1]))

```



## What is mean total number of steps taken per day?
```{r}
# create table with total steps by day
daily_steps <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)

daily_steps_alt1 <- activity_clean %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps))

# this one does not work because it gives dates with only NA total step value 0, so will skew results
# daily_steps_alt2 <- activity %>% 
#   group_by(date) %>% 
#   summarise(steps = sum(steps, na.rm = TRUE))


# Review dataset
head(daily_steps)
summary(daily_steps)

# Average steps per day
mean(daily_steps$steps)

# Median steps per day
median(daily_steps$steps)

# Histogram of total daily steps
ggplot(daily_steps) +
  geom_histogram(aes(x = steps), bins = 20, color = "#000000", fill = "blue") +
  labs(
    title = "Total Daily Steps", 
    x = "Steps", 
    y = "Frequency"
  )


```



## What is the average daily activity pattern?
```{r}

# time series plot of the 5-min intervals x axis and y axis the average steps across all days

avg_steps <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)
head(avg_steps)
summary(avg_steps)

ggplot(avg_steps) +
  geom_line(aes(x = interval, y = steps), color = "blue") +
  labs(
    title = "Average Steps per 5-minute Interval", 
    x = "5-minute Interval", 
    y = "Average Step Count"
  )

# which 5 minute interval contains max number of average steps?
avg_steps$interval[which.max(avg_steps$steps)]
max(avg_steps$steps)

head(avg_steps[order(avg_steps$steps, decreasing = TRUE), ])


```



## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as # NA) The presence of missing days may introduce bias into some calculations or summaries of the data.

Replace missing values with the average value of steps for that 5-minute interval

```{r}

# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA
sum(is.na(activity))


# Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

activity[which(is.na(activity$steps)), ]


# Create a new dataset that is equal to the original dataset but with the missing data filled in.

activity_imp <- activity %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps)) %>% 
  ungroup()


activity_imp
summary(activity_imp)
summary(activity)


# Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

# Create a new dataset with total daily steps with imputed data
daily_steps_imp <- activity_imp %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps))
  
  
head(daily_steps_imp)
head(daily_steps)

summary(daily_steps_imp)
summary(daily_steps)

# comparison chart of steps by day
ggplot(daily_steps_imp, aes(x = date, y = steps))+
  geom_line(color = "red") +
  geom_line(data = daily_steps, color = "gray") +
  labs(
    title = "Total Steps by Day", 
    x = "Date", 
    y = "Total Steps"
  )

# make a histogram of daily steps

ggplot(daily_steps_imp)+
  geom_histogram(aes(x = steps), color = "white", fill = "blue") +
  labs(
    title = "Distribution of Total Daily Steps", 
    x = "Total Daily Steps", 
    y = "Frequency"
  )

# Mean and median number of steps from imputed data
mean(daily_steps_imp$steps)
cat("Mean =", mean(daily_steps_imp$steps), "\n") #\n means line break

median(daily_steps_imp$steps)
cat("Median =", median(daily_steps_imp$steps), "\n")

```




## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

```{r}

# Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.


unique(weekdays(activity_imp$date, abbreviate = TRUE))

activity_imp <- activity_imp %>% 
  mutate(day = weekdays(date, abbreviate = TRUE)) %>% 
  mutate(day_type = ifelse(day %in% c("Mon", "Tue", "Wed", "Thu", "Fri"), "Weekday", "Weekend"))

head(activity_imp)
tail(activity_imp)
str(activity_imp)
activity_imp$day_type <- as.factor(activity_imp$day_type)
str(activity_imp)
summary(activity_imp)

# Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

# create data frame for average steps by interval by day type
avg_steps_imp <- activity_imp %>% 
  group_by(interval, day_type) %>% 
  summarise(avg_steps = mean(steps))

head(avg_steps_imp)

# Make panel plot
ggplot(avg_steps_imp) +
  geom_line(aes(x = interval, y = avg_steps), color = "blue") +
  facet_wrap(~day_type, dir = "v") +
  labs(
    title = "Average Steps per Day by 5-minute Interval", 
    x = "5-minute Interval", 
    y= "Average steps"
  )

```

